---
title: "Newcomb’s speed of light"
subtitle: "Week3-ex3, solution"
output: pdf_document
---

R-template \texttt{ex\_speed\_of\_light.Rmd}.  

Data file \texttt{ex\_speedOfLight.dat}.  


(Here we redo the analysis from page 66 in BDA3.) 

Simon Newcomb conducted experiments on speed of light in 1882. He measured the time required for 
light to travel a certain distance and here we will analyze a data recorded as deviations from $24,\!800$
nanoseconds.  The model used in BDA3 is 

\begin{align}
y_i &\sim N(\mu, \sigma^2) \\
p(\mu,\sigma^2) &\propto \sigma^{-2}
\end{align}

where $y_i$ is the $i$'th measurement, $\mu$ is the mean of the measurement and $\sigma^{2}$ 
the variance of the measurements. Notice that this prior is improper ("uninformative"). 
This corresponds to widely used uniform prior for $\mu$ in the range $(-\infty,\infty)$, and uniform prior for $\log(\sigma)$ (BDA3 pp. 66, 52, and 21). Both priors are improper and 
cannot be found from Stan. 
You can use instead

\begin{align}
p(\mu) &\sim N(0,(10^3)^2)\nonumber\\ 
p(\sigma^2) &\sim \text{Inv-}\chi^2(\nu=4,s^2=1)  \label{eq:Inv-chi_prior}
\end{align}

In this exercise your tasks are the following:

\begin{enumerate}
\item Write a Stan model for the above model and sample from the posterior of the parameters. Report the posterior mean, variance and 95\% central credible interval for $\mu$ and $\sigma^2$.
\item Additionally draw samples from the posterior predictive distribution of hypothetical new measurement $p(\tilde{y}|y)$. Calculate the mean, variance and 95\% quantile of the posterior predictive distribution. 
\item How does the posterior predictive distribution differ from the posterior 
of $\mu$ and Why? 
\item Which parts of the model could be interpreted to correspond to aleatory and epistemic uncertainty? Discuss whether this distinction is useful here. 
\item Instead of Inverse-$\chi^2$ distribution the variance parameter prior has traditionally been defined using Gamma distribution for the precision parameter $\tau=1/\sigma^2$. By using the results in Appendix A of BDA3 derive the analytic form of a Gamma prior for the precision corresponding to the prior \eqref{eq:Inv-chi_prior}. This should be of the form $\text{Gamma}(\alpha,\beta)$, where $\alpha$ and $\beta$ are functions of $\nu$ and $s^2$.
\end{enumerate}

\textbf{Note!} Many common distributions have multiple parameterizations, for which 
reason you need to be careful when interpreting others' works. The 
variance/precision parameter and their priors are notorious for this. The reason 
is mainly historical since different parameterizations correspond to different 
analytical solutions.


\textbf{Grading:} 2 points from correct answer for each of the above steps.
\newpage

## Model answers


Load the needed libraries into R and set options for multicore computer.
```{r}
library(ggplot2)
library(StanHeaders)
library(rstan)
set.seed(123)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

**Part 1 & 2. **

write the model description, set up initial values for 4 chains and sample from the posterior
```{r}
speed_of_light_observations_model = "data {
    int n; //the number of observations
    vector[n] y; //data recorded as deviations from 24800 nanoseconds
  }
parameters {
  real mu; //mean 
  real<lower=0> sigma2; //precision
}
model {
  mu~normal(0,sqrt(10^6));
  sigma2~scaled_inv_chi_square(4,1);
  y ~ normal(mu,sqrt(sigma2));  //or with for loop over the vector elements
  
}
generated quantities {
      // produce samples from the posterior predictive distirbution
      real ytilde;
      ytilde = normal_rng(mu,sqrt(sigma2));
}"

# Load the data and put it into a list format
dataset <- read.table ("ex_speedOfLight.dat", header=TRUE)
#dataset
y <- dataset$y
#hist(y,20)
# set data into named list
data = list(n =nrow(dataset), y =y)

# initialize parameters 
init1 = list(mu = 10, sigma2 = 1)
init2 = list(mu = 20, sigma2 = 10)
init3 = list(mu = 30, sigma2 = 100)
init4 = list(mu = 40, sigma2 = 1000)
inits <- list(init1, init2, init3, init4) 

# Fit a model defined in the Stan modeling language and return the fitted result as an instance of stanfit.
set.seed(123)
post=stan(model_code=speed_of_light_observations_model,data=data,warmup=500,iter=2000,chains=4,thin=1,init=inits,control = list(adapt_delta = 0.8,max_treedepth = 15))
```

Let's then examine the convergence and autocorrelation of the chains.
```{r}
# Check the names of sampled variables
names(post)

#check the convergence visually, plot the sample chains 
plot(post, plotfun= "trace", inc_warmup = TRUE)

# plot autocorrelation function
stan_ac(post,inc_warmup = FALSE, lags = 25)
```

Note, print(post) returns also mean, sd (=square root of variance) and some quantiles (95% central credible interval is defined by 2.5% and 97.5% quantiles) for all parameters. The 95% quantile for $\tilde{y}$ has to be calculated separately though.

```{r}
# print Rhat and other summary statistics
# summary(post)
print(post)
quantile(as.matrix(post, pars =c("ytilde")), 0.95 )
```


**Part 3.**

Note. We had defined the Stan model so that posterior predictive samples for $\tilde{y}$ where generated in the "generated quantities" block. Hence, we can examine them directly using the print(post) command. Let's additionally visualize the posteriors

```{r}
post_samples=as.matrix(post, pars =c("mu","sigma2","ytilde"))
# write.table(post_samples, file="param.txt", row.names=FALSE, col.names=TRUE)

par(mfrow=c(1,3))
hist(y,main='Measured observations',breaks=30)
hist(post_samples[,"mu"], main="Posterior for mu", xlab="mu",breaks=30)
hist(post_samples[,"ytilde"], main="Posterior predictive distribution", xlab="y_pred",breaks=30)
```

From the above histograms we see that the posterior mean of $\mu$ matches that of $\tilde{y}$ whereas the spread of the posterior distribution for $\tilde{y}$ is much wider than that for $\mu$. The reasons are that in the posterior of $\tilde{y}$ 1) we have marginalized over uncertainty in $\mu$ and $\sigma^2$ and 2) we have taken into account the observation error represented by the observation model $\tilde{y}\sim N(\mu,\sigma^2)$. 

Below we demonstrate how to sample from the posterior predictive distribution outside Stan.
```{r}
# Draw samples from the posterior predictive distribution
set.seed(123)
ytilde2 <- rnorm(length(post_samples[,"mu"]), post_samples[,"mu"], sqrt(post_samples[,"sigma2"]))
hist(ytilde2, main="Posterior predictive distribution", xlab="y_pred",breaks=30)
```

**Part 4**

The uncertainty about $\mu$ and $\sigma^2$ is epistemic in nature and it is represented by the prior and posterior distributions. The uncertainty on $y$ or $\tilde{y}$ contains also aleatory uncertainty which is described by the obseration model  $\tilde{y}\sim N(\mu,\sigma^2)$.


**Part 5**

Here the question was to find parameter transformation between Scaled-Inverse-Chi distribution for variance parameter, 

$\sigma^2 \sim \text{Inv-}\chi(\nu=4,s^2=1)$

and Gamma distributions for the precision parameter

$1/\sigma^2 \sim \text{Gamma}(\alpha,\beta)$

From page 579 of BDA3 we get the following relationship. $Inv-Chi(\nu=4,s^2=1)$
 is the same as $\text{Inv-Gamma}(\alpha,\beta)$ with

$\alpha = \frac{\nu}{2}     = 2$

$\beta = \frac{\nu}{2}s^2  = 2$

On the other hand, on page 583 of BDA3 it is told that if $1/\theta$ has a Gamma 
distribution with parameters $\alpha$, $\beta$ then $\theta$ has the inverse-Gamma distribution." 
(And the other way around). Hence, the prior for the precision is 

$1/\sigma^2 \sim \text{Gamma}(2,2)$


# Grading
**Total 10 points.** 2 points from correct answer for each of the above steps. 


